dispRity metrics

### 2016-11-28 doc/dispRity-manual.Rmd

Glossary
--------

@@

Note that we use the `release` branch here which is version 0.2.
For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).

@@

## Loading the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.8) ; axisPhylo(root = 140)
```

@@

```{r}
str(median_centroids)
## That's a more than 4000 lines of output!
```

Therefore this package proposes a specific class of object called `dispRity` objects.
These objects allow to easily use S3 method functions such as `summary.dispRity` (just called as `summary`; see section \[summary\]) or `plot.dispRity` (just called as `plot`; see section \[plot\]).
But also, this allows to use the S3 method for printing `dispRity` objects via `print.dispRity` that allows to summarise the content of the objects similar to the `phylo` class objects (see `ape::print.phylo`).

@@

Note however, that it is always possible to recall the full object using the argument \texttt{all=TRUE}:

```{r, eval=FALSE}
## Displaying the full object
print(median_centroids, all = TRUE)
## But that's 155091 lines!
```

Finally, some utility functions such as `get.dispRity` or `extract.dispRity` allows to access to some specific content of the object:

```{r}
## Extracting some specific series from the disparity object
series_1_and_4 <- get.dispRity(median_centroids, what = c(1,4))
series_1_and_4

## The observed disparity
extract.dispRity(median_centroids)

@@

This package aims to be a modular package where users can personalise some aspects of the package fairly easily.
The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
For example, the `dispRity` function intake a `metric` argument designating how disparity should be calculated.
This argument can take some already implemented functions such as `mean` but also some completely new ones such as the sum divided by length:

@@

## Disparity measured as the mean divided by the length
summary(dispRity(sliced_data, metric = sum.divide.by.length))
## Giving the exact same results!
```

@@

In the following section, different functionalities are detailed. The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
Note however that most of these functions can be used in any order and that some steps can of course be skipped.

One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter *series*).
These series correspond to regions of the ordinated space that we want to compare to each other.

@@

The `cust.series` function allows to create these series according to factors determined by the user.
In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no direct offspring).

@@

## We then can use the customised series function to create the two series
## i.e. the two regions of the ordinated space
crown_stem <- cust.series(BeckLee_mat50, factors)

## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem

@@

Note that here we separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
It is however possible to add more factors with overlapping groups:

@@

1.  discrete time series (or time-binning) using `method = discrete`;

2.  continuous time series (or time-slicing) using `method = continuous`.

@@

1.  defining the boundaries of the time bins (when `method = discrete`);

2.  defining the time slices (when `method = continuous`).

Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a data frame containing the first and last occurrence data (`FAD/LAD`) for taxa that span over a longer time than the tips/nodes age.

@@

```
In this example, the taxa where split inside each timebin according to their age.
Note however that some taxa might span between some time bins and should be included in more than one.
This is possible by providing a table containing the first and last occurrence data:

```{r}
## Generating three time bins containing the data present every 40 Ma, including
## taxa that might span between time bins.
time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
time_bins

@@

1.  `acctran` where the data chosen on each time slice is always the one of the offspring

2.  `deltran` where the data chosen on each time slice is always the one of the descendant

3.  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant

4.  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length

```{r}
## Generating four time slices every 40 million years assuming a gradual
## evolution model
time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
time_slices
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

`boot.matrix` {#rare}
-------------

Once we obtain our different series, we might want to bootstrap and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction.
The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option (see below).

```{r}
## Bootstrapping a matrix
boot.matrix(data = BeckLee_mat50)
## As we can see, the output is also a dispRity object that is summarised
## automatically, and gives information on the data as well as the number of
## bootstraps and the bootstraps methods.
```

@@

1.  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)

2.  `single` where only one random elements is replaced by one other random elements each pseudo-replication

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
It can also be set to a fix `numeric` value (or a set of `numeric` values).
Finally, one last argument, `rm.last.axis` allows to remove a certain amount of dimensions (or axis) for the ordinated space.
This can be logical argument where `FALSE` (default) will not remove any dimension and `TRUE` will remove the last dimensions that bear up to 5% of the total ordinated space’s variance.

@@

```{r}
## Bootstrap and full rarefaction on the crown/stem series
crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)

## Bootstrap on the time binning/slicing series
time_binsBS <- boot.matrix(time_bins, bootstraps = 100)
time_slicesBS <- boot.matrix(time_slices, bootstraps = 100)

## Note that all these objects are of class dispRity
crown_stemBS
time_binsBS
time_slicesBS
```

`dispRity` {#disparity}
----------

@@

In practice, the `dispRity` function intakes one or more functions as a definition of disparity. Several of these functions will be already implemented in other packages (such as `stats::median`, `base::sum`, etc.); some others are implemented in this package (listed in `?dispRity.metric`) and finally some others will be defined by the users.
The `make.metric` function is designed to help users create and test their own disparity metric functions. In practice, the use of these metrics in the `dispRity` function, is pretty easy:

@@

Note that these functions do not directly output the disparity values but only the summary of the `dispRity` objects.
To display the results, see the section \[summary\].

In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the `dispRity` objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
In this example we are going to define disparity as being the median distance between each element and each dimensions centroids.

@@

`summary`
---------

@@

## If on happens to forgot what was calculated in the summarised object,
## it is possible to recall the different steps details using recall:
summary(disp_time_binsBS, recall = TRUE)

@@

`plot`
------

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")

@@

## Same plot but with the number of elements on the right and using the
## "polygon" plot type.
plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
## Rarefied version of the same plot (20 elements per series)
plot(disp_crown_stemBS, type = "polygon", elements = TRUE, rarefaction = 20)
```

@@

-   `pairwise` (default): to compare each series pairwise

-   `referential`: to compare each series to the first one

-   `sequential`: to compare each series to the following one

-   `all`: to compare all the series together (like in analysis of variance)

>>>>

Glossary <a name="glossary"></a>
--------

@@

Note that we use the `release` branch here which is version `0.2`.
For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).

@@

## Loading the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.7) ; axisPhylo(root = 140)
```

@@

```{r, eval = FALSE}
str(median_centroids)
## That's a more than 5000 lines of output!
```

Therefore this package proposes a specific class of object called `dispRity` objects.
These objects allow to easily use S3 method functions such as `summary.dispRity` (just called as `summary`; see [summary](#summary) section) or `plot.dispRity` (just called as `plot`; see [plot](#plot) section).
But also, this allows to use the S3 method for printing `dispRity` objects via `print.dispRity` that allows to summarise the content of the objects similar to the `phylo` class objects (see `ape::print.phylo`).
The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).

@@

Note however, that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:

```{r, eval=FALSE}
## Displaying the full object
print(median_centroids, all = TRUE)
## But that's more than 155000 lines!
```

Some utility functions such as `sort.dispRity`, `get.dispRity` or `extract.dispRity` allows to access or reorder to some specific content of the object:

```{r}
## Reordering the series (from the present to the past)
sort.dispRity(median_centroids, decreasing = TRUE)

## Extracting some specific series from the disparity object
get.dispRity(median_centroids, what = c(1,4))

## Extracting the observed disparity
extract.dispRity(median_centroids, observed = TRUE)

@@

This package aims to be a *modular package*, that is, it is designed to be fairly easily modified by users.
The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
For example, the `dispRity` function intake a `metric` argument designating how disparity should be calculated.
This argument can take some already implemented functions such as `mean` but it can also intake some completely new and revolutionary ones such as the sum divided by length:

@@

## Disparity measured as the mean divided by the length
summary(dispRity(sliced_data, metric = sum.divide.by.length))
## That is a bit like reinventing the wheel...
```

@@

In the following section, different functionalities are detailed.
The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
Note however that most of these functions can be used in any order and that some steps can of course be skipped and that the package contains many more additional functions!

One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *series*, see [glossary](#glossary)).
These series correspond to regions of the ordinated space that we want to compare to each other.

@@

The `cust.series` function allows to create these series according to factors determined by the user.
In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (i.e. all the living mammals and their direct ancestor) and the stem mammals (i.e. all the mammals that have no related living descendants).

@@

## We then can use the customised series function to create the two series
## i.e. the two regions of the ordinated space
crown_stem <- cust.series(BeckLee_mat50, factors)
```

The results of this function (like many others presented below) output a `dispRity` object:

```{r}
crown_stem

@@

Note that here we are separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
It is however possible to add more factors with overlapping groups:

@@

 *  discrete time series (or time-binning) using `method = discrete`;

 *  continuous time series (or time-slicing) using `method = continuous`.

@@

 *  defining the boundaries of the time bins (when `method = discrete`);

 *  defining the time slices (when `method = continuous`).

Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a data frame containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age.

@@

In this example, the taxa where split inside each time-bin according to their age.
Note however that some taxa might span between some time bins and should be included in more than one.
This is possible by providing a table containing the first and last occurrence data:

```{r}
## Generating three time bins containing the data present every 40 Ma, including
## taxa that might span between time bins.
(time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))

@@

 *  `acctran` where the data chosen on each time slice is always the one of the offspring

 *  `deltran` where the data chosen on each time slice is always the one of the descendant

 *  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant

 *  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length

```{r}
## Generating four time slices every 40 million years assuming a gradual
## evolutionary model
(time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages))
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

`boot.matrix`
-------------

Once we obtain our different series, it is possible to bootstrap and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction.
The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option (see below).

```{r}
## Bootstrapping a matrix
boot.matrix(data = BeckLee_mat50)
```

@@

 *  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)

 *  `single` where only one random elements is replaced by one other random elements each pseudo-replication

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
It can also be set to a fix `numeric` value (or a set of `numeric` values).
Finally, one last argument, `rm.last.axis` allows to remove a certain amount of dimensions (or axis) for the ordinated space.
This can be logical argument where `FALSE` (default) will not remove any dimension and `TRUE` will remove the last dimensions that bear up to 5% of the total ordinated space’s variance.
Note that it can also intake a `numerical` value designating the proportion of axis to keep.

@@

```{r}
## Bootstrap and full rarefaction on the crown/stem series
(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))

## Bootstrap on the time binning/slicing series
(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
```

`dispRity`
----------

@@

In practice, the `dispRity` function intakes one or more functions as a definition of disparity. Several of these functions will be already implemented in other packages (such as `stats::median`, `base::sum`, etc.); some others are implemented in this package (listed in `?dispRity.metric`) and finally some others can be defined by the users.
The `make.metric` function is designed to help users create and test their own disparity metric functions.
In practice, the use of these metrics in the `dispRity` function, is pretty easy:

@@

These functions do not directly output the disparity values but only the summary of the `dispRity` objects.
To display the results, see the section [summary](#summary).

In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the `dispRity` objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
Here we are going to define disparity as being the median distance between each element and each dimensions centroids.

@@

`summary` <a name="summary"></a>
---------

@@

## If one happens to forgot what was calculated in the summarised object,
## it is possible to recall the different steps details using recall:
summary(disp_time_binsBS, recall = TRUE)

@@

`plot` <a name="plot"></a>
------

@@

```{r, fig.width=6, fig.height=6}
## Graphical parameters
op <- par(mfrow = c(2,2), bty = "n")

@@

## Same plot but with the number of elements on the right and using the
## "polygon" plot type.
plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
## Rarefied version of the same plot (20 elements per series)
plot(disp_crown_stemBS, type = "polygon", rarefaction = 20)
```

@@

 *   `pairwise` (default): to compare each series pairwise

 *  `referential`: to compare each series to the first one

 *  `sequential`: to compare each series to the following one

 *  `all`: to compare all the series together (like in analysis of variance)

### 2016-11-28 doc/dispRity-metrics.Rmd

# A two dimensional disparity
all_variances <- summary(dispRity(BeckLee_mat50, metric = variances))

# Transforming the two dimensional one in a one dimensional one
sum_variances2 <- summary(dispRity(BeckLee_mat50, metric = sum))

>>>>

# A two dimensional disparity
all_variances <- dispRity(BeckLee_mat50, metric = variances)

# Transforming the two dimensional one in a one dimensional one
sum_variances2 <- summary(dispRity(all_variances, metric = sum))
